---
title: Energetic Consequences of Sonar Exposure for Cetaceans
author:
  - name: Max F Czapanskiy
    email: maxczap@stanford.edu
    affiliation: "1"
    footnote: Corresponding author
  - name: Matthew S Savoca
    affiliation: "1"
  - name: Will T Gough
    affiliation: "1"
  - name: Jeremy A Goldbogen
    affiliation: "1"
address:
  - code: "1"
    address: Department of Biology, Hopkins Marine Station, Stanford University, 120 Ocean View Boulevard, Pacific Grove, CA 93950, USA
abstract: |
  The sub-lethal effects of sonar exposure for cetaceans is not well enough understood to predict population consequences. To better inform conservation planning, we developed a model to predict the energetic costs of sonar avoidance behavior. When avoiding sonar sources, cetaceans may cease foraging and/or flee the area. We estimate the potential energy intake lost to foraging cessation and the additional locomotor costs from increased swim speeds.

  Although medium-sized cetaceans, especially beaked whales, make up the majority of mass strandings associated with naval sonar, our results indicate sub-lethal effects cause the most harm to large baleen whales. A severe response by a Cuvier's beaked whale (6.5 hour foraging cessation and 30 minutes of elevated swim speeds at 4.5 m/s) cost 49% of its daily basal metabolic requirements. Conversely, a moderate response by a blue whale (1 hour foraging cessation and 5 minutes of elevated swim speeds at 2.5 m/s) cost more than nine times its daily basal metabolic requirements. These results may be used to reduce harm to endangered cetacean species. 

journal: "JEB? RSPB?"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
  - \usepackage{setspace}
  - \usepackage{lineno}
  - \usepackage{threeparttable}
  - \usepackage{float}
---

\doublespacing
\linenumbers

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.height = 3, fig.width = 5)
```

Introduction
============

Naval exercises involving sonar have been linked to mass strandings of cetaceans worldwide since at least the 1980's [@simmonds_whales_1991; @frantzis_does_1998; @england_joint_2001; @jepson_gas-bubble_2003]. However, the population consequences of sub-lethal effects are not as well understood. Controlled exposure experiments (CEEs) show that behavioral responses may include cessation of feeding and/or fleeing the sonar source at elevated speed [@tyack_beaked_2011; @deruiter_first_2013; @goldbogen_blue_2013; @friedlaender_prey-mediated_2016; @deruiter_multivariate_2017; @kvadsheim_avoidance_2017; @southall_behavioral_2019; @wensveen_northern_2019]. Quantitatively linking these behaviors to demographics requires an understanding of the impacts on individuals' health [@pirotta_understanding_2018]. The mechanism addressed here is reduced energy stores due to lost foraging opportunities and increased locomotor costs. 

The two extant clades of cetaceans differ in their feeding styles. Toothed whales (odontocetes) are raptorial feeders and locate prey using echolocation [ref?]. Baleen whales (mysticetes) are filter feeders and capture prey by lunging [@goldbogen_scaling_2012] or continuous ram filter feeding [@werth_baleen_2016]. These feeding styles have profound effects on feeding rate, energy per feeding event, dive depth, and body size. Odontocetes feed at higher rates on smaller prey and most larger odontocetes must dive to extreme depths to find sufficient prey [ref?].  Lunge-feeding mysticetes (rorquals) engulf enormous quantities of prey-laden water, increasing the energy intake per feeding event but limiting dive depth and duration [@goldbogen_scaling_2012]. 

Feeding rates and the energy obtained per feeding event can be empirically measured with multi-sensor tags, active acoustics, and stomach contents [jeremy's paper?]. The rapid echolocation clicking (buzzes) preceeding odontocete prey capture events have a kinematic and acoustic signature that register on tag sensors [ref?]. Similarly, rorqual lunges can be identified from an increase in speed followed by a rapid deceleration, usually associated with high pitch and roll angles [@cade_kinematic_2016]. Simultaneous prey mapping with tagging efforts using active acoustics have measured the biomass density of fish and krill schools targeted by rorquals. The sizes of squid beaks and fish otoliths *(?)* in odontocete stomach samples are a record of prey sizes and energy per feeding event.

Empirical estimates of cetacean metabolic rates are logistically challenging for smaller species and infeasible for larger species. Oxygen consumption has been measured for captive odontocetes trained to swim under a metabolic hood using open-flow respirometry, showing that mass-specific stroke costs are largely size invariant [@williams_physiology_1993; @williams_swimming_2017]. Whether these metabolic estimates scale to larger odontocetes or mysticetes is unknown, so other methods of estimating energy expediture include breathing rates and hydrodynamic models [@sumich_swimming_1983; @potvin_metabolic_2012].

Animals swim efficiently by cruising at 1-2 m/s and maintaining a Strouhal number of 0.25 - 0.3 [@katsufumi_stroke_2007; @rohr_strouhal_2004]. The Strouhal number is a dimensionless quantity $St = \frac{Af}{U}$ where $A$ is stroke amplitude, $f$ is stroke frequency, and $U$ is swimming speed. Cetacean stroke amplitudes are approximately one fifth body length [ref?] so there is a linear relationship between swimming speed and stroke frequency for animals of a given body size [gough?].

Methods
=======

We considered the potential energy intake of lost feeding opportunities and additional energy expenditure from elevated swimming speeds in modeling the energetic consequences of sonar exposure. The model takes the form:

\begin{align}
E_{sonar} = P_{in} \times t_d + P_{out}(U_f) \times t_f
\end{align}

Where $E_{sonar}$ is the energy cost of sonar exposure, $P_{in}$ is consumption power (i.e. rate of energy intake) during undisturbed foraging, $t_d$ is the time displaced from foraging, $P_{out}$ is flight power (i.e. increased rate of locomotor costs), $U_f$ is the animal's speed during flight, and $t_f$ is the flight time. $P_{in}$ and $P_{out}$ are species-specific values and $t_d$, $U_f$, and $t_f$ are dependent on the individual's behavioral response to sonar exposure.

## Consumption power ($P_{in}$)

The rate of energy intake is the product of feeding rate ($r_f$) and prey energy per feeding event ($E_p$):

\begin{align}
P_{in} = r_f \times E_p
\end{align}

$r_f$ was calculated as the lunge rate for rorquals and buzz rate for odontocetes using tag sensors. $E_p$ was derived using active acoustics (rorquals) and stomach contents (odontocetes) [ref?].

## Flight power ($P_{out}$)

The locomotor costs associated with fleeing the sonar source is the energetic cost of swimming at $U_f$ relative to the cruising swim speed. Using the relationship between stroke frequency and swimming speed and a scaling relationship for mass-specific stroke costs, we calculate $P_{out}$ as:

\begin{align}
P_{out}(U_f) = (f_s(U_f) - f_s(U_c)) \times C_L \times m
\end{align}

Where $f_s$ is a function relating stroke frequency to swimming speed, $U_f$ and $U_c$ are the swimming speeds during the flight response and cruising, $C_L$ is the mass-specific locomotor cost of a stroke, and $m$ is the animal's mass. Equation (2) assumes, during the flight response, cetaceans increase swimming speed and stroke frequency, but the mass-specific locomotor cost of a stroke remains the same regardless of speed. Although $C_L$ increases with swimming speed, the scaling relationships do not hold for larger cetaceans [@williams_swimming_2017]. To be conservative, we use the $C_L$ scaling relationship for cruising speeds, estimated as $C_L = 1.46 + 0.0005m$ in $J \cdot stroke^{-1} \cdot kg^{-1}$. We chose 1.5 m/s for $U_c$ based on size-invariant scaling of cruising speeds [@katsufumi_stroke_2007]. Assuming cetaceans maintain a Strouhal number of 0.3 and stroke amplitude is one fifth body length, stroking frequency as a function of body length is:

\begin{align}
S_t &= \frac{Af}{U} \notag\\
f &= \frac{1.5U}{L}
\end{align}

## Case studies
We analyzed tag data from four controlled exposure experiments (CEEs) and applied the $E_{sonar}$ model to estimate energetic costs of observed behavioral responses to sonar exposure [BRS ref?]. As part of an on-going behavioral response study to naval sonar, in 2011-2015 *(???)* Blainville's beaked whales (*Mesoplodon densirostris*), Cuvier's beaked whales (*Ziphius cavirostris*), northern minke whales (*Balaenoptera acutorostrata*), and blue whales (*B. musculus*) were tracked with multi-sensor tags and exposed to mid-frequency active sonar in Southern California. Time displaced from foraging ($t_d$), time in flight ($t_f$), and speed of flight ($U_f$) were selected based on tag data. These values should be considered realistic scenarios, but not the typical or most common response. Behavioral responses to sonar exposure are highly variable and seem to depend on received sound level, distance to source, foraging behavior, and habituation [@southall_behavioral_2019; @wensveen_northern_2019].

```{r setup}
library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
library(lmodel2)
library(ggsci)

# Load data
# Morphological data
abbr_binom = function(binom) {
  paste(str_sub(binom, 1, 1), 
        str_extract(binom, " .*"), 
        sep = ".")
}
morphologies <- read_csv("../data/foragestats_combined_ko2.csv") %>% 
  mutate(Species = str_replace(Species, "_", " ")) %>% 
  group_by(Species) %>% 
  summarize(Length_m = first(Body_length_m),
            Mass_kg = first(Body_mass_kg)) %>% 
  mutate(Clade = ifelse(str_detect(Species, ".*ptera.*"),
                        "Mysticete",
                        "Odontocete"),
         Family = recode(Species,
                         `Balaenoptera bonaerensis` = "Balaenopteridae",
                         `Balaenoptera musculus` = "Balaenopteridae",
                         `Balaenoptera physalus` = "Balaenopteridae",
                         `Berardius bairdii` = "Ziphiidae",
                         `Globicephala macrorhynchus` = "Delphinidae",
                         `Globicephala melas` = "Delphinidae",
                         `Grampus griseus` = "Delphinidae",
                         `Megaptera novaeangliae` = "Balaenopteridae",
                         `Mesoplodon densirostris` = "Ziphiidae",
                         `Orcinus orca` = "Delphinidae",
                         `Phocoena phocoena` = "Phocoenidae",
                         `Physeter macrocephalus` = "Physeteridae",
                         `Ziphius cavirostris` = "Ziphiidae")) %>% 
  # binomial is a factor ordered by species length with an abbreviated label
  arrange(Length_m) %>% 
  mutate(binomial = factor(Species, 
                           levels = unique(Species)))
binom_levels <- levels(morphologies$binomial)

# Prey data
prey_data <- read_csv("../data/Cetacea model output BOUT_EXTANT_v2.csv") %>% 
  select(-(X13:X25)) %>% 
  mutate(Species = recode(Species,
                          bonarensis = "bonaerensis",
                          Phocaena = "phocoena"),
         Genus = recode(Genus,
                        Physter = "Physeter",
                        Phocoaena = "Phocoena"),
         binomial = factor(paste(Genus, Species),
                           levels = binom_levels)) %>% 
  filter(`MR exponent` == 0.45)  # data duplicated for multiple MR exponent values

# Rorqual feeding rates
rorqual_data <- read_csv("../data/lunge_rates_from_Paolo.csv") %>%
  filter(species != "be") %>% 
  mutate(duration_h = `deployment-time_secs` / 60 / 60,
         binomial = factor(recode(species,
                                  ba = "Balaenoptera bonaerensis",
                                  bp = "Balaenoptera physalus",
                                  bw = "Balaenoptera musculus",
                                  mn = "Megaptera novaeangliae"),
                           levels = binom_levels),
         lunge_h = total_lunges / duration_h)

# Odontocete feeding rates
odontocete_data <- read_csv("../data/foragestats_combined_ko2.csv") %>%
  mutate(binomial = factor(str_replace(Species, "_", " "),
                           levels = binom_levels)) %>% 
  filter(taxa == "O")

# Labels for logarithmic scales
log_labels <- trans_format("log10", math_format(10^.x))
```

Results
=======

## Consumption power ($P_{in}$)
Feeding rates were derived from tag data and tended to decrease with body size. Rorqual feeding rates ranged from 10.7 to 44.1 lunges/hour and odontocete feeding rates ranged from 2.5 to 96.8 buzzes/hour (Table \ref{rf_tbl}).

```{r results="asis"}
# Rorqual lunge rates
lunge_rate <- rorqual_data %>% 
  filter(duration_h >= 24) %>%
  group_by(Species = binomial) %>% 
  summarize(N = n(),
            `$r_f$` = round(mean(lunge_h), digits = 1)) %>% 
  mutate(Group = "Rorqual")

# Odontocete buzz rates
buzz_rate <- odontocete_data %>%  
  group_by(Species = binomial) %>% 
  summarize(N = n(),
            `$r_f$` = (total_buzz_count / total_duration_h) %>% 
              mean(na.rm = TRUE) %>% 
              round(digits = 1)) %>% 
  mutate(Group = "Odontocete")

# Feeding rate table
rbind(buzz_rate, lunge_rate) %>% 
  mutate(Species = sprintf("\\textit{%s}", Species)) %>%
  select(Group, Species, `$r_f$`, N) %>% 
  kable("latex", 
        booktabs = TRUE, 
        escape = FALSE,
        caption = "Cetacean feeding rates \\label{rf_tbl}",
        linesep = "") %>%
  footnote(general = c("$r_f$ is lunges/hour (rorquals) or buzzes/hour (odontocetes)", 
                       "Species ordered by size within groups."),
           escape = FALSE,
           threeparttable = TRUE)
```

Prey energy per feeding event was empirically derived from acoustic backscatter (mysticetes) and stomach samples (odontocetes). Filter feeders consumed the most energy per feeding event. Generally, delphinids targeted more energy-rich prey than Ziphids (Fig. \ref{Ep_fig}).

Modeled consumption power covered four orders of magnitude: from 2.4e3 kJ/hr (*P. phocoena*) to 1.4e7 kJ/hr (*B. musculus*) (Fig. \ref{Pin_fig}).

## Flight power ($P_{out}$)

## Case studies

Figures
=======
```{r fig.cap = "Energy per feeding event. Note log scale on y-axis. \\label{Ep_fig}", fig.height=4.5}
prey_data %>% 
  uncount(weights = Percent) %>% 
  mutate(binomial = fct_relabel(binomial, abbr_binom)) %>% 
  ggplot(aes(binomial, `Energy (kJ)`, color = Family)) +
  geom_boxplot() +
  scale_y_log10(name = "Energy per feeding event (kJ)",
                labels = log_labels) +
  scale_color_aaas() +
  expand_limits(y = 1) +
  theme_classic(base_size = 12) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(hjust = 1,
                                   angle = 30,
                                   face = "italic"),
        legend.title = element_blank(),
        legend.position = "bottom",
        legend.direction = "horizontal",
        legend.text = element_text(size = 8))
```

```{r fig.cap = "Modeled consumption power ($P_{in}$). \\label{Pin_fig}"}
Ep_tbl <- prey_data %>%
  uncount(weights = Percent) %>% 
  group_by(binomial) %>% 
  summarize(Ep_kJ = mean(`Energy (kJ)`))

rf_tbl <- rorqual_data %>% 
  filter(duration_h >= 24) %>%
  group_by(binomial) %>% 
  summarize(rf_h = round(mean(lunge_h), 1)) %>%
  rbind(odontocete_data %>%  
          group_by(binomial) %>% 
          summarize(rf_h = round(mean(total_buzz_count / total_duration_h, na.rm = TRUE), 1)))

Pin_tbl <- inner_join(Ep_tbl, rf_tbl, by = "binomial") %>% 
  mutate(Pin_kJ_h = Ep_kJ * rf_h) %>% 
  left_join(select(morphologies, binomial, Family))

Pin_tbl %>% 
  mutate(binomial = fct_relabel(binomial, abbr_binom)) %>% 
  ggplot(aes(binomial, Pin_kJ_h, fill = Family)) +
  geom_col() +
  scale_y_log10(name = expression(P[`in`] ~~ (kJ/h)),
                labels = log_labels) +
  scale_fill_aaas() +
  expand_limits(y = 1) +
  theme_classic(base_size = ) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(hjust = 1,
                                   angle = 30,
                                   face = "italic"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.direction = "horizontal",
        legend.text = element_text(size = 8))
```

References {#references .unnumbered}
==========
