---
title: "Supporting Information"
subtitle: "Large baleen and small toothed whales face greatest energetic consequences from sonar disturbance"
author: 
 - "Max F. Czapanskiy"
 - "Matthew S. Savoca"
 - "William T. Gough"
 - "Paolo S. Segre"
 - "Danuta M. Wisniewska"
 - "David E. Cade"
 - "Jeremy A. Goldbogen"
date: "3/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(cowplot)
library(knitr)
library(kableExtra)
library(scales)
library(tidyverse)
```

## Data

The sonar response energetic model uses three types of data: feeding rates, prey quality, and morphology. The feeding rate, morphology, and odontocete prey quality data were previously published by Goldbogen et al. (2019) <doi:10.1126/science.aax9044>. See supplement "Description of how rorqual prey resources were calculated" for description of rorqual prey quality data. 

```{r data}
prey <- readRDS("prey_tbl.RDS")
feeding <- readRDS("feeding_tbl.RDS")
morphology <- readRDS("morpho_tbl.RDS")

cbf_palette <- c(
  # blue
  "Delphinidae and Phocoenidae" = rgb(0, 114, 178, maxColorValue = 255),  
  # vermillion
  "Physeteridae and Ziphiidae" = rgb(213, 94, 0, maxColorValue = 255),
  # bluish green
  "Balaenopteridae" = rgb(0, 158, 115, maxColorValue = 255)
) 

morphology %>% 
  mutate(species = cell_spec(species, "latex", italic = TRUE)) %>% 
  select(Species = species, Family = family, Clade = clade, 
         `Length (m)` = length_m, `Mass (kg)` = mass_kg, 
         Abbreviation = abbr) %>% 
  kable("latex", booktabs = TRUE, escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down"))
```

For the model, feeding rate ($r_f$) and energy from prey per feeding event ($E_p$) distributions are empirical and lognormal, respectively. The following table combines the three data sources and provides quantile functions for $r_f$ and $E_p$ for each species.

```{r combine_data}
# Creates a quantile function based on the log-mean and log-sd of each species'
#   prey distribution
prey_fun <- prey %>% 
  mutate(q_Ep_kJ = map2(meanlnEp_lnkJ, 
                        sdlnEp_lnkJ, 
                        ~ function(p) qlnorm(p, .x, .y)))
# Creates an empirical quantile function for each species
feeding_fun <- feeding %>% 
  group_by(species) %>% 
  summarize(rf_h = list(rf_h)) %>% 
  mutate(q_rf_h = map(rf_h, ~ function(p) quantile(.x, p)))
# Combines the three data sources
species_data <- morphology %>% 
  left_join(prey_fun, by = "species") %>% 
  left_join(feeding_fun, by = "species")
```

## Baseline rate of energy acquisition

The baseline rate of energy acquisition ($\frac{d}{dt}E_a$) was modeled as the product of $r_f$ and $E_p$. The resulting distribution for blue whales (*Balaenoptera musculus*) is presented in the main manuscript (Fig. 1). The distributions for all species are below.

```{r ddtEa}
# Adapted from https://stackoverflow.com/a/45867076
label_sci <- function(breaks) {
  ifelse(
    breaks == 0, 
    "0", 
    parse(text = gsub("[+]", "", gsub("e", " %*% 10^", scientific(breaks))))
  )
}

# Plot the lognormal distribution of energy from prey
Ep_plot <- function(q_Ep_kJ, meanlnEp_lnkJ, sdlnEp_lnkJ) {
  tibble(x = q_Ep_kJ(c(0.001, 0.99))) %>% 
    ggplot(aes(x)) +
    stat_function(
      fun = dlnorm,
      args = list(meanlog = meanlnEp_lnkJ,
                  sdlog = sdlnEp_lnkJ),
      n = 1e3
    ) +
    expand_limits(x = 0) +
    scale_x_continuous(expression(italic(E[p]) ~ (kJ)), 
                       labels = label_sci) +
    theme_classic(base_size = 8) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(),
          axis.title.x = element_text(size = 8))
}

# Plot the empirical distribution of feeding rates
rf_plot <- function(q_rf_h) {
  tibble(x = q_rf_h(seq(0, 1, length.out = 1000))) %>% 
    ggplot(aes(x)) + 
    geom_histogram(bins = 20, 
                   boundary = 0, 
                   fill = "light gray", 
                   color = "black",
                   size = 0.2) +
    scale_x_continuous(expression(italic(r[f]) ~ ("hr"^{-1}))) +
    theme_classic(base_size = 8) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(),
          axis.title.x = element_text(size = 8))
}

# Plot the baseline rate of energy acquisition by sampling Ep and rf
ddtEa_plot <- function(species, q_Ep_kJ, q_rf_h, caption = TRUE) {
  ddtEa <- tibble(p_Ep = runif(1e3),
                  p_rf = runif(1e3)) %>% 
    mutate(Ep = q_Ep_kJ(p_Ep),
           rf = q_rf_h(p_rf),
           ddtEa = Ep * rf)
  
  ddtEa_caption <- if (caption) {
    element_text(face = "italic")
  } else {
    element_blank()
  }
  
  ddtEa %>% 
    filter(ddtEa < quantile(ddtEa, 0.99)) %>% 
    ggplot(aes(ddtEa)) +
    geom_histogram(bins = 30,
                   boundary = 0, 
                   fill = "light gray", 
                   color = "black") +
    geom_vline(aes(xintercept = mean(ddtEa)),
               linetype = "dashed") +
    scale_x_continuous(expression(italic(frac(d, dt) * E[a]) ~~ ("kJ" ~ "hr"^{-1})),
                       labels = label_sci) +
    labs(caption = species) +
    theme_classic(base_size = 12) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(),
          axis.title.x = element_text(size = 8),
          plot.caption = ddtEa_caption) 
}

# Combine plots (Ep and rf as insets in d/dt Ea)
arrange_ddtEa <- function(species, q_Ep_kJ, meanlnEp_lnkJ, sdlnEp_lnkJ, q_rf_h, 
                          caption = TRUE, ...) {
  Ep <- Ep_plot(q_Ep_kJ, meanlnEp_lnkJ, sdlnEp_lnkJ)
  rf <- rf_plot(q_rf_h)
  ddtEa <- ddtEa_plot(species, q_Ep_kJ, q_rf_h, caption)
  
  # Figure out where the insets should go
  scales <- layer_scales(ddtEa)
  xlim <- scales$x$get_limits()
  ylim <- scales$y$get_limits()
  grid_to_data <- function(xmin, xmax, ymin, ymax) {
    c(xmin = xlim[1] + xmin * (xlim[2] - xlim[1]),
      xmax = xlim[1] + xmax * (xlim[2] - xlim[1]),
      ymin = ylim[1] + ymin * (ylim[2] - ylim[1]),
      ymax = ylim[1] + ymax * (ylim[2] - ylim[1]))
  }
  Ep_coord <- grid_to_data(0.1, 0.535, 0.45, 0.95)
  rf_coord <- grid_to_data(0.555, 0.99, 0.45, 0.95)
  ddtEa +
    annotation_custom(ggplotGrob(Ep), 
                      xmin = Ep_coord["xmin"], xmax = Ep_coord["xmax"],
                      ymin = Ep_coord["ymin"], ymax = Ep_coord["ymax"]) +
    annotation_custom(ggplotGrob(rf), 
                      xmin = rf_coord["xmin"], xmax = rf_coord["xmax"],
                      ymin = rf_coord["ymin"], ymax = rf_coord["ymax"])
}

# Figure 1 of main text
species_data %>% 
  filter(abbr == "Bm") %>% 
  pmap(arrange_ddtEa, caption = FALSE)

ddtEa_plots <- species_data %>% 
  select(species, q_Ep_kJ, meanlnEp_lnkJ, sdlnEp_lnkJ, q_rf_h) %>% 
  pmap(arrange_ddtEa)
walk(ddtEa_plots, print)
```

$r_f$ and mass-specific $E_p$ across species (Figure 2 in main text).

```{r fig2}
fig2data <- species_data %>% 
  mutate(
    rf_mean = map_dbl(rf_h, mean),
    rf_1q = map_dbl(rf_h, quantile, p = 0.25),
    rf_3q = map_dbl(rf_h, quantile, p = 0.75),
    Epmass_mean = exp(meanlnEp_lnkJ) / mass_kg,
    Epmass_1q = map2_dbl(meanlnEp_lnkJ, sdlnEp_lnkJ, ~ qlnorm(0.25, .x, .y)) / mass_kg,
    Epmass_3q = map2_dbl(meanlnEp_lnkJ, sdlnEp_lnkJ, ~ qlnorm(0.75, .x, .y)) / mass_kg,
    abbr = factor(abbr, levels = .$abbr)
  )
# Fig 2a
fig2a <- ggplot(fig2data, aes(abbr, rf_mean)) +
  geom_pointrange(aes(ymin = rf_1q, ymax = rf_3q, 
                      color = group, shape = group)) +
  scale_color_manual(values = cbf_palette) +
  labs(x = "", y = expression(italic(r[f])~~(hr^{-1}))) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
# Fig 2b
fig2b <- ggplot(fig2data, aes(abbr, Epmass_mean)) +
  geom_pointrange(aes(ymin = Epmass_1q, ymax = Epmass_3q, 
                      color = group, shape = group)) +
  scale_color_manual(values = cbf_palette) +
  labs(x = "", y = expression("Mass-specific"~~italic(E[p])~~(kJ~kg^{-1}))) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
# Fig 2
plot_grid(fig2a, fig2b, labels = "AUTO")
```

## Energy expenditure due to elevated locomotion

Elevated locomotion is defined here as a speed increase ($U_f$) for a duration ($t_f$). The energetic cost associated with this behavior was modeled as the product of the increase in stroking frequency ($\Delta f$ in Hz), the mass-specific locomotor cost ($C_L$ in kJ stroke^-1^ kg^-1^), and the individual's mass. To incorporate uncertainty, both $\Delta f$ and $C_L$ were treated as gamma distributed variables with a mean equal to the best estimate and a shape parameter of 4. The following figure shows the estimated increase in mass-specific energy expenditure when $U_f$ is 5 m s^-1^. 

```{r ddtEe}
# Predict the change in fluking frequency (deltaff) and mass-specific locomotor
# costs (CL)
U_cruise_ms <- 1.5
U_flight_ms <- 5
# Change in fluking frequency in strokes per hour
ff_fun <- function(U, L, La = 0.2, St = 0.3) {
  St * U / L / La * 3600
}
# Mass-specific locomotor cost in kJ / kg / stroke
CL_fun <- function(m) (1.46 + 0.0005 * m) / 1000

# Estimate the quantile of mass-specific ddtEe (product of two gammas)
q_ddtEe <- function(p, deltaff_shape, deltaff_scale, CL_shape, CL_scale) {
  pmap_dbl(
    list(deltaff_shape, deltaff_scale, CL_shape, CL_scale),
    function(deltaff_shape, deltaff_scale, CL_shape, CL_scale) {
      tibble(
        deltaff = rgamma(1e6, shape = deltaff_shape, scale = deltaff_scale),
        CL = rgamma(1e6, shape = CL_shape, scale = CL_scale)
      ) %>% 
        mutate(ddtEe = deltaff * CL) %>% 
        pull(ddtEe) %>% 
        quantile(p)
    })
}

# Plot median and IQR of mass-specific ddtEe
species_data %>% 
  mutate(
    ff_cruise = ff_fun(U_cruise_ms, length_m),
    ff_flight = ff_fun(U_flight_ms, length_m),
    mean_deltaff = ff_flight - ff_cruise,
    deltaff_shape = 4,
    deltaff_scale = mean_deltaff / deltaff_shape,
    mean_CL = CL_fun(mass_kg),
    CL_shape = 4,
    CL_scale = mean_CL / CL_shape,
    ddtEe_1q = q_ddtEe(0.25, deltaff_shape, deltaff_scale, CL_shape, CL_scale),
    ddtEe_med = q_ddtEe(0.5, deltaff_shape, deltaff_scale, CL_shape, CL_scale),
    ddtEe_3q = q_ddtEe(0.75, deltaff_shape, deltaff_scale, CL_shape, CL_scale),
    species = factor(species, levels = species[order(length_m)])
  ) %>% 
  ggplot(aes(x = species, y = ddtEe_med)) +
  geom_pointrange(aes(ymin = ddtEe_1q, ymax = ddtEe_3q)) +
  coord_flip() +
  labs(y = expression("Mass-specific " ~~ italic(frac(d, dt) * E[e]) ~~ ("kJ" ~ "hr"^{-1} ~ "kg"^{-1}))) +
  theme_classic(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic"),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_blank()) 
```

## Sensitivity analysis

We tested the model's sensitivity to energy acquisition ($E_p$, $r_f$) and expenditure ($\Delta f$, $C_L$) parameters in two behavioral scenarios. The scenarios were chosen to emphasize increased energy expenditure ($t_d$ = 1 hour, $t_f$ = 0.5 hours, $U_f$ = 5 m s^-1^) or lost consumption ($t_d$ = 4 hours, $t_f$ = 0.25 hours, $U_f$ = 3.5 m s^-1^). 

```{r sensitivity1}
# Energetic cost of sonar exposure
Esonar_fun <- function(td_hr, tf_hr, rf, Ep, delta_ff, CL, mass) {
  ddtEa <- Ep * rf # kJ * hr-1 = kJ hr-1
  ddtEe <- delta_ff * CL * mass # strokes hr-1 * kJ kg-1 stroke-1 * kg = kJ hr-1
  ddtEa * td_hr + ddtEe * tf_hr # kJ hr-1 * hr + kJ hr-1 * hr = kJ
}
# Run a scenario
run_scenario <- function(species, scenario,
                         td_hr, tf_hr, Uf_ms,
                         q_rf_h, meanlnEp_lnkJ, sdlnEp_lnkJ, 
                         length_m, mass_kg, ...) {
  shape_arg <- 4
  delta_ff <- ff_fun(Uf_ms, length_m) - ff_fun(1.5, length_m)
  CL <- CL_fun(mass_kg)
  
  # Parameter list
  param <- c("rf_h", "Ep_kJ", "delta_ff", "CL")
  # Quantile functions
  q <- list(rf_h = q_rf_h,
            Ep_kJ = qlnorm,
            delta_ff = qgamma,
            CL = qgamma)
  # Parameter arguments
  q_arg <- list(
    rf_h = list(),
    Ep_kJ = list(meanlog = meanlnEp_lnkJ, 
                 sdlog = sdlnEp_lnkJ),
    delta_ff = list(shape = shape_arg, scale = delta_ff / shape_arg),
    CL = list(shape = shape_arg, scale = CL / shape_arg)
  )
  # Generate Latin hypercube samples and run model
  pse::LHS(model = NULL, param, 5e2, q, q_arg)$data %>% 
    mutate(
      esonar_kJ = Esonar_fun(td_hr, tf_hr, rf_h, Ep_kJ, delta_ff, CL, mass_kg)
    ) %>% 
    {cbind(tibble(species, scenario, td_hr, tf_hr, Uf_ms), .)}
}

# Run both behavioral scenarios
sensitivity_tbl <- 
  tribble(~scenario,     ~td_hr, ~tf_hr, ~Uf_ms,
          "flight",      1,      0.5,    5,
          "consumption", 4,      0.25,   3.5) %>% 
  crossing(species_data) %>% 
  pmap_dfr(run_scenario)
```

Model sensitivity to each parameter was quantified as the coefficient in the linear model $z(E) \sim z(r_f) + z(E_p) + z(\Delta f) + z(C_L)$ where $z()$ is the z-score. Results by guild are presented in the main text (Fig. 3). The following are the results for each species.

```{r sensitivity2}
# Normalize parameters by z-score
zscore <- function(x) (x - mean(x)) / sd(x)
sensitivity_coef <- sensitivity_tbl %>% 
  group_by(species, scenario) %>% 
  mutate_at(vars(esonar_kJ, rf_h, Ep_kJ, delta_ff, CL), zscore) %>% 
  group_modify(function(data, keys) {
    # Fit linear model
    esonar_lm <- lm(esonar_kJ ~ rf_h + Ep_kJ + delta_ff + CL, data = data) 
    # Extract coefficients and confidence intervals
    esonar_coef <- coef(esonar_lm)
    esonar_ci <- as_tibble(confint(esonar_lm, level = 0.95),
                           rownames = "param")
    colnames(esonar_ci)[2:3] <- c("ci_min", "ci_max")
    
    cbind(esonar_coef, esonar_ci) %>% 
      select(param, esonar_coef, ci_min, ci_max) %>% 
      # drop intercept
      slice(-1)
  }) %>% 
  ungroup() %>% 
  mutate(
    param = factor(
      param, 
      levels = c("rf_h", "Ep_kJ", "delta_ff", "CL"),
      labels = c("r[f]", "E[p]", "Delta*f", "C[L]")
    ),
    species = factor(
      species,
      levels = species_data$species[order(species_data$length_m)]
    )
  )

# Plot sensitivity coefficients and confidence intervals for all species
param_lbls <- parse(text = levels(sensitivity_coef$param))
ggplot(sensitivity_coef, aes(species, esonar_coef, color = param)) +
  geom_pointrange(aes(ymin = ci_min, ymax = ci_max), 
                  position = position_dodge(1),
                  size = 0.25) +
  scale_color_viridis_d(labels = param_lbls) +
  coord_flip() +
  facet_wrap(~ scenario) +
  labs(y = "Sensitivity") +
  theme_classic(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic"),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(hjust = 0),
        legend.title = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank()) 

# Sensitivity coefficients and CIs by guid
sensitivity_guilds <- sensitivity_tbl %>% 
  left_join(select(species_data, species, group), by = "species") %>% 
  mutate(group = factor(group, 
                        levels = c("Balaenopteridae",
                                   "Physeteridae and Ziphiidae",
                                   "Delphinidae and Phocoenidae"))) %>% 
  group_by(group, scenario) %>% 
  mutate_at(vars(esonar_kJ, rf_h, Ep_kJ, delta_ff, CL), zscore) %>% 
  group_modify(function(data, keys) {
    # Fit linear model
    esonar_lm <- lm(esonar_kJ ~ rf_h + Ep_kJ + delta_ff + CL, data = data) 
    # Extract coefficients and confidence intervals
    esonar_coef <- coef(esonar_lm)
    esonar_ci <- as_tibble(confint(esonar_lm, level = 0.95),
                           rownames = "param")
    colnames(esonar_ci)[2:3] <- c("ci_min", "ci_max")
    
    cbind(esonar_coef, esonar_ci) %>% 
      select(param, esonar_coef, ci_min, ci_max) %>% 
      # drop intercept
      slice(-1)
  }) %>% 
  ungroup() %>% 
  mutate(
    param = factor(
      param, 
      levels = c("rf_h", "Ep_kJ", "delta_ff", "CL"),
      labels = sprintf("italic(%s)", 
                       c("r[f]", "E[p]", "Delta*f", "C[L]"))
    )
  )
fig3 <- sensitivity_guilds %>% 
  group_by(scenario) %>% 
  group_map(
    function(data, ...) {
      ggplot(data,
             aes(x = param, y = esonar_coef, 
                 ymin = ci_min, ymax = ci_max, 
                 color = group)) +
        geom_pointrange(position = position_dodge(0.6), 
                        size = 1, 
                        fatten = 0.75) +
        coord_flip() +
        scale_x_discrete(
          labels = parse(text = levels(sensitivity_guilds$param))
        ) +
        scale_y_continuous(
          "Sensitivity",
          breaks = seq(0, 1, by = 0.2),
          labels = seq(0, 1, by = 0.2)
        ) +
        scale_color_manual(values = cbf_palette) +
        expand_limits(y = c(0, 1)) +
        theme_classic(base_size = 18) +
        theme(axis.title.y = element_blank(),
              legend.position = "none",
              strip.background = element_blank())
    }
  ) %>% 
  plot_grid(plotlist = ., labels = "AUTO")

# Table of sensitivity coefficients and CIs
abbr_binom <- function(binom) {
  paste(str_sub(binom, 1, 1), 
        str_extract(binom, " .*"), 
        sep = ".")
}
coef_tbl <- sensitivity_coef %>% 
  mutate(abbr = abbr_binom(species),
         esonar_fmt = sprintf("%0.2f (%0.2f - %0.2f)", 
                              esonar_coef, ci_min, ci_max),
         species = as.character(species)) %>% 
  left_join(select(species_data, species, length_m), by = "species") %>% 
  arrange(length_m) %>% 
  select(scenario, abbr, param, esonar_fmt) %>% 
  pivot_wider(names_from = param, values_from = esonar_fmt) %>% 
  arrange(scenario) %>% 
  select(-scenario)
colnames(coef_tbl) <- c("", "$r_f$", "$E_p$", "$\\Delta f$", "$C_L$")

coef_tbl %>%
  kable("latex", booktabs = TRUE, escape = FALSE) %>% 
  kable_styling(latex_options = c("scale_down")) %>% 
  row_spec(0, align = "c") %>% 
  pack_rows("Lost consumption scenario", 1, 11) %>%
  pack_rows("Increased expenditure scenario", 12, 22)
```

## Critical cessation threshold

Cross-species comparisons of energetic costs are complicated by the range of body sizes 
across cetaceans. In absolute terms, a loss of 10,000 kJ would be extreme for a harbor porpoise but likely insubstantial to a blue whale. To account for these differences, we used metabolic scaling to approximately estimate the daily energy budget for each species and modeled the duration of a feeding cessation that would exceed the daily energy budget ($t_{crit}$). The results presented in the main text (Fig. 4) are for metabolic scaling according to Kleiber (1975) and Maresh (2014) and an FMR:BMR ratio ($\beta$) of 3. For comparison, the following calculations include $\beta$ of 2.5 and 5.

```{r tcrit}
# Estimate FMR from mass and beta (Kleiber and Maresh curves)
fmr_fun <- function(mass, beta, bmr) {
  daily_bmr <- if (bmr == "Kleiber") {
    293.1 * mass ^ 0.75
  } else if (bmr == "Maresh") {
    581 * mass ^ 0.68
  } else {
    stop("Unspecified BMR calculation")
  } 
  beta * daily_bmr
}

# Estimate tcrit once
tcrit_fun <- function(Uf_ms, tf_hr, beta, bmr, mass_kg, length_m,
                      rf_h, meanlnEp_lnkJ, sdlnEp_lnkJ, ...) {
  # Mean delta_ff and CL
  delta_ff <- ff_fun(Uf_ms, length_m) - ff_fun(1.5, length_m)
  CL <- CL_fun(mass_kg)
  
  # Simulate 14 days of feeding cessation and find the fmr crossing-point
  fmr <- fmr_fun(mass_kg, beta, bmr)
  cumulative_Esonar <- tibble(hour = seq(1, 14 * 24)) %>% 
    mutate(
      rf = sample(rf_h, nrow(.), replace = TRUE),
      Ep = rlnorm(nrow(.), meanlog = meanlnEp_lnkJ, sdlog = sdlnEp_lnkJ),
      delta_ff = rgamma(nrow(.), shape = 4, scale = delta_ff / 4),
      CL = rgamma(nrow(.), shape = 4, scale = CL / 4),
      Esonar = Esonar_fun(td_hr = 1, tf_hr, rf, Ep, delta_ff, CL, mass_kg),
      cum_Esonar = cumsum(Esonar)
    )
  
  approx(x = c(0, cumulative_Esonar$cum_Esonar),
         y = c(0, cumulative_Esonar$hour),
         xout = fmr,
         ties = mean)$y
}

# Flight scenarios for tcrit calculations
tcrit_tbl <- tribble(
  ~scenario,        ~tf_hr,  ~Uf_ms,
  "no_flight",      0,       1.5,
  "mild_flight",    5 / 60,  2.5,
  "strong_flight",  15 / 60, 3.5,
  "extreme_flight", 30 / 60, 5
  ) %>% 
  mutate(scenario = factor(scenario,
                           levels = c("no_flight", 
                                      "mild_flight",
                                      "strong_flight",
                                      "extreme_flight"),
                           labels = c("No flight",
                                      "Mild flight",
                                      "Strong flight",
                                      "Extreme flight"))) %>% 
  # Estimate tcrit for all scenarios 1000 times
  crossing(species_data,
           beta = c(2.5, 3, 5),
           bmr = factor(c("Kleiber", "Maresh")),
           i = 1:1000) %>% 
  mutate(tcrit = pmap_dbl(., tcrit_fun))

# Summarize tcrit results
tcrit_summ <- tcrit_tbl %>% 
  group_by(abbr, scenario, bmr, beta) %>% 
  summarize(tcrit_mean = mean(tcrit),
            tcrit_1q = quantile(tcrit, 0.25),
            tcrit_3q = quantile(tcrit, 0.75)) %>% 
  ungroup() %>% 
  mutate(abbr = factor(abbr, levels = species_data$abbr))
```

```{r tcrit_results}
# Figure 4 from main text
filter(tcrit_summ, beta == 3) %>%
  ggplot(aes(abbr,
             tcrit_mean,
             color = scenario,
             shape = bmr)) +
  geom_pointrange(aes(ymin = tcrit_1q,
                      ymax = tcrit_3q),
                  fatten = 2,
                  position = position_dodge(width = 0.6)) +
  # scale_x_discrete(labels = morphologies$abbr) +
  scale_y_continuous(
    expression(italic(t[crit])),
    breaks = c(0.5, 1, 4, 12, 24, 48, 72),
    minor_breaks = NULL,
    labels = c("30 min", "1 hour", "4 hours", "12 hours",
               "1 day", "2 days", "3 days"),
    trans = "log2"
  ) +
  scale_color_brewer(palette = "RdYlBu", direction = -1) +
  labs(y = "Feeding cessation") +
  theme_classic(base_size = 12) +
  theme(axis.title.x = element_blank(),
        legend.box.margin = margin(),
        legend.justification = c(0, 0),
        legend.position = c(0.11, 0.08),
        legend.title = element_blank()) +
  guides(color = guide_legend(ncol = 2), shape = guide_legend(ncol = 2))

# Table of tcrit results
tcrit_table <- tcrit_summ %>%
  mutate(tcrit_fmt = sprintf("%.3g (%.3g - %.3g)",
                             tcrit_mean, tcrit_1q, tcrit_3q)) %>%
  select(-(tcrit_mean:tcrit_3q)) %>%
  pivot_wider(names_from = c("bmr", "beta"), values_from = tcrit_fmt)
colnames(tcrit_table) <- c(
  "", "",
  "$t_{crit} (hr), \\beta = 2.5$", 
  "$t_{crit} (hr), \\beta = 3$", 
  "$t_{crit} (hr), \\beta = 5$",
  "$t_{crit} (hr), \\beta = 2.5$", 
  "$t_{crit} (hr), \\beta = 3$", 
  "$t_{crit} (hr), \\beta = 5$"
)
kable(tcrit_table, "latex", booktabs = TRUE, escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down")) %>% 
  add_header_above(c(" " = 2, "Kleiber" = 3, "Maresh" = 3)) %>% 
  collapse_rows(columns = 1, latex_hline = "major", valign = "top") 
```
